# Comparing the means of two groups

## Learning objectives {-}

By the end of this module you will be able to:

- Decide whether to use an independent samples t-test or a paired t-test to compare two groups for a continuous outcome variable;
- Conduct and interpret the results from an independent samples t-test;
- Describe the assumptions of an independent samples t-test;
- Conduct and interpret the results from a paired t-test;
- Describe the assumptions of a paired t-test;
- Conduct an independent samples t-test and a paired t-test in Stata;
- Report results and provide a concise summary of the findings of statistical analyses.

## Readings {-}

@kirkwood_sterne01a; Sections 7.1 to 7.5.

@bland15b; Section 10.3.

@acock10; Section 7.7, 7.8.

@juul_frydenberg14; Juul S and Frydenberg M, An Introduction to Stata for Health Researchers, 4th Edition (2014): Section 11.5.

## Introduction

In Module 4, a one-sample t-test was used for comparing a mean value with a hypothesised value. In this module, we show how to compare the mean values of two groups for which the outcome variable is normally distributed. In health research, we often want to compare the mean value of a measurement between two groups in an observational study or between a control and intervention group in an experimental study. For example, in an observational study, we may want to compare cholesterol levels in people who are normal weight to the levels in people who are overweight. In a clinical trial, we may want to compare cholesterol levels in people who have been randomised to a dietary modification or to usual care.

From the decision tree presented in the Appendix, we can see that if we have a continuous outcome measure and two categorical groups that are not related, i.e. a binary exposure measurement, the test for such data is an independent samples t-test. The test is also sometimes called a 2-sample t-test.

However, in research, data are often 'paired' or 'matched', that is the two data points are related to one another. This occurs when measurements are taken:

- From each participant on two occasions, e.g. at baseline and follow-up in an experimental study or in a longitudinal cohort study;
- From related people, e.g. a mother and daughter or a child and their sibling;
- From related sites in the same person, e.g. from both limbs, eyes or kidneys;
- From matched participants e.g. in a matched case-control study;
- In cross-over clinical trials where the patient receives both drugs, often in random order.

An independent samples t-test cannot be used for analysing paired or matched data because the assumption that the two groups are independent is violated. Treating paired or matched measurements as independent samples would artificially inflate the sample size and lead to inaccurate analyses and biased P values. When the data are related in a paired or matched way and the outcome is continuous, a paired t-test is the appropriate statistic to use if the data are normally distributed.

## Independent samples t-test

An independent samples t-test is a parametric test that is used to assess whether the mean values of two groups are different from one another. Thus, the test is used to assess whether two mean values are similar enough to have come from the same population or whether the difference between them is so large that the two groups can be considered to have come from separate populations with different characteristics.

The null hypothesis is that the mean values of the two groups are not different, that is:

H<sub>0</sub>: (Mean2 -- Mean1) = 0

Rejecting the null hypothesis using an independent samples t-test indicates that the difference between the means of the two groups is large in relation to the variation in the samples and is unlikely to be due to chance or to sampling variation.

### Assumptions for an independent samples t-test

The assumptions that must be met before an independent samples t-test can be used are:

- The two groups are independent
- The measurements are independent
- The outcome variable must be continuous and must be normally distributed in each group
- The variance in the two groups is similar (homogenous)

The first two assumptions are determined by the study design. The two samples must be independent, i.e. if a person is in one group then they cannot be included in the other group, and the measurements within a sample must be independent, i.e. each person must be included in their group once only.

The third assumption of normality is important although t-tests are robust to some degree of non-normality as long as there are no influential outliers and, more importantly, if the sample size is large. We examined how to assess normality in Module 2. If the data are not normally distributed, it may be possible to transform them using a mathematical function such as a logarithmic transformation. If not, then we may need to use non-parametric tests. This is examined in Module 9.

The final assumption is homogeneity of variance between the groups. This can be verified by checking that the standard deviation (square root of the variance) of each group is similar. If the variances are different, then Welch's t-test, an alternative version of the t-test can be used.

### Worked Example

In an observational study of a random sample of 100 full term babies from the community, birth weight and gender were measured. There were 44 male babies and 56 female babies in the sample. The research question asked whether there was a difference in birth weights between boys and girls. The two groups are independent of each other and therefore an independent samples t-test can be used to test the null hypothesis that there is no difference in weight between the genders.

Some preliminary descriptive statistics of the distribution of the variable of interest in each group should always be obtained before a t-test is undertaken to ensure that the assumptions are met. Box plots and histograms are ideal for this. Histrograms and box plots of the data obtained in Stata using **Graphics > Box plot** is shown in Figure 5.1. The dataset Example_5.1.dta is available on Moodle.

<!-- INSERT FIGURE -->

The plots show that the data are approximately normally distributed: the histograms are relatively bell shaped and symmetric, and the boxes are fairly symmetrical, there are no outliers as indicated by dots, and the spread is similar in both groups as the similar length of the whiskers suggesting that the variance is homogenous.

We can obtain statistics using the summarize command by gender with the detail option to check the data (e.g. skewness, plausibility of the minimum and maximum values).

```{r, echo=TRUE, eval=FALSE, highlight=FALSE}
. by gender, sort: summarize birthweight , detail

---------------------------------------------------------------------
-> gender = Female

                         Birthweight
-------------------------------------------------------------
      Percentiles      Smallest
 1%         2.95           2.95
 5%         3.03           2.97
10%         3.14           3.03       Obs                  56
25%        3.325           3.07       Sum of Wgt.          56

50%         3.53                      Mean           3.587411
                        Largest       Std. Dev.      .3629788
75%         3.88            4.2
90%         4.15            4.2       Variance       .1317536
95%          4.2            4.2       Skewness       .2453238
99%         4.25           4.25       Kurtosis       1.962126

---------------------------------------------------------------------
-> gender = Male

                         Birthweight
-------------------------------------------------------------
      Percentiles      Smallest
 1%         2.75           2.75
 5%         2.82           2.79
10%         2.85           2.82       Obs                  44
25%         3.15           2.85       Sum of Wgt.          44

50%         3.43                      Mean           3.421364
                        Largest       Std. Dev.      .3536165
75%        3.635           3.94
90%          3.9           3.97       Variance       .1250446
95%         3.97           4.06       Skewness      -.0895932
99%          4.1            4.1       Kurtosis       2.325761
```

The table shows that girls have a mean weight of 3.59 kg (SD 0.36) and boys have a mean weight of 3.42 kg (SD 0.35) with females being heavier than males. The variabilities of birth weight, as indicated by the standard deviations, are similar.

### Conducting and interpreting an independent samples t-test

An independent samples t-test provides us with a t statistic from which we can compute a P value. The computation of the t statistic is as follows:

$$t = \frac{{\overline{x}}_{1} - {\overline{x}}_{2}}{SE({\overline{x}}_{1} - {\overline{x}}_{2})}$$

with *n~1~ + n~2~* -- 2 degrees of freedom.

Given that the standard error is estimated from the variance, the t value is an estimate of how different the mean values are compared to their variability. Thus, the t value will become larger as the difference in means increases with respect to the variability.

In Stata, both the t and P values are provided. If the t-value falls outside a critical range, the P value will be small and we can reject the null hypothesis of no difference between the groups.

Output 5.3 shows the Stata results of the example dataset obtained using **Statistics - Summaries, tables, and tests - Classical tests of hypotheses - t test (mean-comparison test)** and choosing the two-sample test in the ttest dialog box.

#### Output 5.3: Independent samples t-test results from Stata

```{r, echo=TRUE, eval=FALSE, highlight=FALSE}
. ttest birthweight, by(gender)

Two-sample t test with equal variances
------------------------------------------------------------------------------
   Group |     Obs        Mean    Std. Err.   Std. Dev.   [95% Conf. Interval]
---------+--------------------------------------------------------------------
  Female |      56    3.587411    .0485051    .3629788    3.490204    3.684617
    Male |      44    3.421364    .0533097    .3536165    3.313854    3.528873
---------+--------------------------------------------------------------------
combined |     100     3.51435    .0366567    .3665666    3.441615    3.587085
---------+--------------------------------------------------------------------
    diff |            .1660471    .0723027                .0225648    .3095293
------------------------------------------------------------------------------
    diff = mean(Female) - mean(Male)                              t =   2.2966
Ho: diff = 0                                     degrees of freedom =       98

    Ha: diff < 0                 Ha: diff != 0                 Ha: diff > 0
 Pr(T < t) = 0.9881         Pr(|T| > |t|) = 0.0238          Pr(T > t) = 0.0119
```

The output table reports the mean, standard deviation, 95% confidence interval etc of weights of the two groups separately as well as that of their difference. It shows the mean difference in weights between the genders is 0.17 kg (95% CI 0.02, 0.31). We are 95% confident that the true mean difference lies between 0.02 and 0.31, this interval does not contain the null value of 0.

Stata reports 3 different P-values. Among them the middle one reports the P-value for a two-sided test evaluating the null hypothesis "Difference =0" and is our desired test. The test has a t-value of 2.297 with 98 degrees of freedom, and a two-sided P value of 0.024 which is less than 0.05 and is statistically significant. Thus, we can reject the null hypothesis of no difference in weights between the genders.

## Paired t-tests

If the outcome of interest is the difference in the continuously distributed outcome measurement between each pair or between each case and its matched control, i.e. the within-pair differences a paired t-test is used. In effect, a paired t-test is used to assess whether the mean of the differences between the two related measurements is significantly different from zero. In this sense, a paired t-test is very closely aligned with a one sample t-test.

When using a paired t-test, the variation between the pairs of measurements is the most important statistic and the variation between the participants, which is critical for the interpretation of a two-sample t-test, is of little interest.

For related measurements, the data for each pair of values must be entered on the same row of the spreadsheet. Thus, the number of rows in the data sheet is the number of participants or the number of participant-pairs when cases and controls are matched. Thus, the effective sample size is the total number of pairs and not the total number of measurements.

### Assumptions for a paired t-test

For a paired samples t-test, it is not important to test whether the measurements are normally distributed for each of the time points in the two matched samples, but it is important to test whether the differences between the two measurements are normally distributed.

The assumptions for a paired t-test are:

-   the outcome variable is continuous
-   the differences between the pair of the measurements are normally distributed

If the assumptions for a paired t-test cannot be met, a non-parametric equivalent is a more appropriate test to use (Module 9).

### Computing a paired t-test

The null hypothesis for using a paired t-test is as follows:

H~0~: Mean (Measurement1 -- Measurement2) = 0

To compute a t-value, the size of the mean difference between the two measurements is compared to the standard error of the paired differences, i.e.

$$t = \frac{\overline{d}}{SE(\overline{d})}$$

with *n*--1 degrees of freedom, where *n* is the number of pairs.

Because the standard error becomes smaller as the sample size becomes larger, the t-value increases as the sample size increases for the same mean difference.

### Worked Example 5.2

A total of 107 people were recruited into an experimental trial to assess whether ankle blood pressure measured in two different sites would be the same. For each person, systolic blood pressure (SBP) was measured in two sites: dorsalis pedis and tibialis posterior.

The dataset Example_5.2.dta is available on Moodle. First, we need to compute the pairwise difference between SBP measured in the two sites in Stata using the generate command. This is shown in the Stata manual at the end of this module (Checking the assumptions for a Paired t-test) and in the Foundations module. The distribution of the difference between SBP measured in dorsalis pedis and tibialis posterior is shown in Figure 5.2. It approximates a normal distribution and therefore a paired t-test can be used.

#### Figure 5.2: Distribution of differences in ankle SBP between two sites of 107 participants

[INSERT FIGURE]

The paired t-test is performed using the ttest command in Stata (see the Stata Notes section for details). We specify the data is paired with sbp_dp as First variable and sbp_tp as the Second variable. Output 5.4 shows the summary statistics for both sites. From this we can see that the mean SBP is very similar in the two sites.

#### Output 5.4: Paired t-test results from Stata
```{r, echo=TRUE, eval=FALSE, highlight=FALSE}
Paired t test
------------------------------------------------------------------------------
Variable |     Obs        Mean    Std. Err.   Std. Dev.   [95% Conf. Interval]
---------+--------------------------------------------------------------------
  sbp_dp |     107     116.729    3.460296    35.79358    109.8686    123.5893
  sbp_tp |     107    117.9907    3.431356    35.49422    111.1877    124.7937
---------+--------------------------------------------------------------------
    diff |     107   -1.261682    1.311368    13.56489   -3.861596    1.338232
------------------------------------------------------------------------------
     mean(diff) = mean(sbp_dp - sbp_tp)                           t =  -0.9621
 Ho: mean(diff) = 0                              degrees of freedom =      106

 Ha: mean(diff) < 0           Ha: mean(diff) != 0           Ha: mean(diff) > 0
 Pr(T < t) = 0.1691         Pr(|T| > |t|) = 0.3382          Pr(T > t) = 0.8309
```

The next line for "diff" shows the statistics for the mean of within-pair difference. It indicates that average SBP measured in dorsalis pedis is 116.7 mmHg and that in tibialis posterior is 118.0 mmHg. The difference is −1.26 (95% CI: −3.86 to 1.34).

The t-value of −0.96 yields a two-sided P-value of 0.34 (under Ha: mean(diff) ≠ 0) confirms that these data provide no evidence against the null hypothesis, and conclude that the SBP measured in the two sites are not different.

As with any statistical test, it is important to decide what mean difference between measurements would be considered clinically important in addition to considering statistical significance.
