# **5** R notes {-}

## Setup

For this module, we will use the `tidyverse`, `haven`, `skimr` and `moments` packages:

```{r mod5-r-setup}
library(tidyverse)
library(haven)
library(skimr)
library(moments)
```

## Checking data for the independent samples t-test

We import the `Example_5.1.dta` dataset using the `haven` package, and inspect it using the `skim` function within the `skimr` package:

```{r mod5-r-import-babies}
babies <- read_stata("data/examples/Example_5.1.dta") %>% 
  mutate(gender = factor(gender, levels=c(1,2), labels=c("Male", "Female")))

# skim(babies)
```

### Producing histograms and boxplots by a second variable

To obtain histograms for separate variables in `ggplot2`, we defined a variable as a *facet*.

```{r mod5-r-plot-babies}
ggplot(babies, aes(x = birthweight)) +
  geom_histogram(fill = "white", colour = "black", binwidth=0.25) +
  facet_wrap(gender ~ .)

ggplot(babies, aes(x = birthweight)) +
  geom_density() +
  facet_wrap(gender ~ .)

ggplot(babies, aes(y = birthweight, x=gender)) +
  geom_boxplot()
```

This looks like a cool HTML way to hide solutions:

<details>
  <summary>Solution:

```{r eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```
</details>

### Producing split summary statistics

To produce summary statistics for a continuous variable, split by a second binary categorical variable, as in Worked Example 5.1 using `Example_5.1.dta`, go to **Statistics > Summaries, tables, and tests > Summary and descriptive statistics > Summary statistics** . In the **summarize** dialog box, select `birthweight` as the **Variable** in the **Main** tab of the dialog box. Next go to the **by/if/in** tab, tick **Repeat command by groups** and select `gender` as the **Variable that define groups** as shown below.
 
